Mixture Models for Single-Cell Assays with Applications to Vaccine Studies

### 2013-06-03 MIMOSA.tex
cut cut cut 23 pages

%\subsection{Prior}
As in the one-biomarker case, we share information across subjects using an exchangeable prior on the unknown proportions. The beta priors in  \ref{ss:priors} are replaced by Dirichlet distributions.
%\begin{align*}
%(\mathbf{p}^{(u)}_{i}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\\\nonumber
 %(\mathbf{p}^{(s)}_{i}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}^{(s)}) \quad \text{and}\quad (\mathbf{p}^{(u)}_{i}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\end{align*}
The indicator variable $z_i$ is defined as in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$. As in the beta-binomial case, both an EM and MCMC algorithms can be used for parameter estimation. When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}^{\{u,s\}}$ and $w$ as for the beta-binomial model.

%\subsection{Parameter estimation}
Again, to simplify the estimation problem, we make use of the marginal likelihoods that can be obtained in closed forms (see Appendix C of the Supplementary Material). 
%For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
%\begin{align*}
%\mathrm{L}_0(\boldsymbol{\alpha}^{(u)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) &= \frac{ \mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \cdot \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!},
%\end{align*}
%where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$. Similarly the marginal likelihood for the alternative model is given by
%\[
%\mathrm{L}_1(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}_{i}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) =\frac{\mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}) \mathrm{B}(\boldsymbol{\alpha}^{(s)}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(s)})\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!}.
%\]
The estimation procedures (both EM and MCMC based) for the Dirichlet--multinomial distribution are the same as for the beta-binomial model except that the number of parameters to estimate is larger. We initialize the $z_i$ in the EM algorithm with the positivity calls from the multivariate Fisher's exact test. In our experience, the performance of the EM algorithm greatly deteriorates for $K>3$, and is more dependent on the initial values and can fail to converge in many instances. Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large. More details are given in Appendix C of the Supplementary Material.

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we applied our multivariate MIMOSA model for two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example, $K=2$, and we have four possible combinations.
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the $\log_2$ counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the sum of the counts in the 3 different positive subsets (Figure~\ref{fig:polyfunctionality} A,B). The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all possible polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ or GZMK$+$), would not be appropriate in this case, since changes in the counts of these different cell populations occur in both directions. The multivariate MIMOSA model tests all polyfunctional cell subpopulations simultaneously, and detects significant differences between stimulated and unstimulated conditions in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, dark labels). In contrast, marginalizing over these cell populations, no difference is apparent in any of the samples (Figure~\ref{fig:polyfunctionality} B). Testing all combinations simultaneously is an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

@@

Since the Fluidigm data set has a limited number of observations (100 cells and 16 samples), we could not look at more than two biomarkers at once. Therefore, we performed simulations in eight dimensions to assess the power of the multivariate MIMOSA  model compared to Fisher's exact test and the likelihood ratio test on the resulting 2x8 tables (Supplementary Figure 6 A,B). These results show that multivariate MIMOSA has significantly increased power to detect true differences in multivariate data, even with small counts and small effect sizes, and the model better fits the data than the competing standard approaches tested (Supplementary Figure 6 B).

@@

\section{Discussion}
\label{s:discussion}
The development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important as single-cell assays become more routine and  sequencing at the single-cell level eventually becomes practical~\citep{Ramskold:2012gj}. Current data analysis approaches such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small. Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton2004,Smyth:2005iy,Gottardo:2006}. In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.

The MIMOSA model presented here uses a mixture model framework of beta-binomial or Dirichlet-multinomial distributions to model counts in experimental subjects across multiple conditions (\textit{i.e.}, vaccine responders and non-responders). Information is shared across responders and non-responders through exchangeable priors, increasing the power to detect true differences between treatment and control conditions compared to Fisher's exact test, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 5). The univariate MIMOSA model based on the Beta-Binomial distribution allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the matched unstimulated sample. This has proven to be useful for ICS data where stimulation induced changes are expected to be one-sided. Although in some cases the FDR is not well controlled by any method tested (Supplementary Figure 1A), volcano plots of effect size vs posterior probability of response show these false positives are pre--vaccine subjects exhibiting response  to stimulation (Supplementary Figure 7). Such variability is occasionally expected and may be biological or a technical assay problem.

Our model is fit to dichotomized data  (cells are either positive or negative). While this is the standard approach to ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information. Likewise for single-cell gene expression data, most methods have focused on differences in the continuous part of the signal while using the discrete on/off nature for pre-filtering~\citep{Flatz:2011jb}. We have shown elsewhere that changes in both the frequency of expression and expression level are informative~{McDavid2012}. The ability of MIMOSA to identify stimulation-specific expression patterns in the Fluidigm data demonstrates not only the broader utility of the method, but importantly, further demonstrates that biologically relevant signal is present in the proportion of cells expressing each gene under different conditions (Figure~\ref{fig:fluidigm}). 
Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative sets, counted and compared across different conditions. Extensions of the MIMSOA model, incorporating the continuous MFI or gene expression level are warranted in the future.

@@

Detecting differences in poly-functional cell populations (\textit{i.e.}, identifying changes in cell populations that co-express multiple proteins, cytokines, or genes) is important in immunology, since it allows the identification of more homogeneous cell populations~\citep{Milush:2009bz}.
In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells detected.
Hence, if a specific cell subset expressing multiple biomarkers is being differentially expressed, differential expression based on the marginal cell counts should also be detected.
As such, identifying poly-functional cytokine profiles from ICS data can be done in an iterative way.
First, univariate tests on marginal populations are performed, and then specific cell subsets expressing the positive biomarkers detected are tested.
However, this iterative (univariate) approach might not be satisfactory due to the large number of possible combinations that need to be tested, and a multivariate approach might be preferable.
In that case,  as others have pointed out, in order to have the most power to detect a true difference, the statistical test should be selected taking into account only the cytokine combinations of interest~\citep{Nason:2006dx}.

For two--sided changes, as with the Fluidigm data, changes in poly-functional cell populations are not always detectable when looking at the marginal populations (Figure~\ref{fig:polyfunctionality} A-C).
In this case, the use of multivariate models, as our Dirichlet-multinomial model, will become important to detect differences.
Here, we have shown that MIMOSA has higher sensitivity and specificity than the competing methods to identify true differences between conditions in multivariate count data (Figure~\ref{fig:polyfunctionality} A, and Supplementary Figure 6 A,B), and the model generally provides a better fit to the single-cell assay count data obtained from studies with these types of experimental designs (Supplementary Figure 6 B).
Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
In the case of more than two biomarkers, the number of parameters for our model is $2^{K+1}+1$, which is large even for moderate values of K. As an example, we would need both, a large number of subjects and a large number of events (cells) collected, to properly estimate the 33 parameters for $K=4$. A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters.
For example, one could assume that the hyper-parameters are constant across biomarker combinations, \textit{i.e.}, $\alpha^{\{s,u\}}_{k}=\alpha^{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain biomarkers more than others.
More exploratory work will need to be done in this area once high dimensional single-cell level data with large number of samples become available.

>>>>

%\subsection{Prior}
As in the one-biomarker case, we share information across subjects using an exchangeable prior on the unknown proportions. The beta priors in  \ref{ss:priors} are replaced by Dirichlet distributions.
%\begin{align*}
%(\mathbf{p}^{(u)}_{i}|z_i=0) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\\\nonumber
 %(\mathbf{p}^{(s)}_{i}|z_i=1) \sim \mathrm{Dir}(\boldsymbol{\alpha}^{(s)}) \quad \text{and}\quad (\mathbf{p}^{(u)}_{i}|z_i=1) &\sim \mathrm{Dir}(\boldsymbol{\alpha}^{(u)}),\end{align*}
The indicator variable $z_i$ is defined as in Section \ref{ss:priors}, \textit{i.e.}, $z_i\sim\mathrm{Be}(w)$. As in the beta-binomial case, both EM and MCMC algorithms can be used for parameter estimation. When using a fully Bayesian approach via MCMC, we use the same priors for $\boldsymbol{\alpha}^{\{u,s\}}$ and $w$ as for the beta-binomial model.

%\subsection{Parameter estimation}
We again make use of the marginal likelihoods in closed form (see Appendix C of the Supplementary Material). 
%For the null component, the marginal likelihood $\mathrm{L}_0$ is given by,
%\begin{align*}
%\mathrm{L}_0(\boldsymbol{\alpha}^{(u)}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) &= \frac{ \mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \cdot \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!},
%\end{align*}
%where $\mathrm{B}$ is the $2^K$-dimensional Beta function defined as $\mathrm{B}(\boldsymbol{\alpha})=\prod_k\Gamma(\alpha_k)/\Gamma(\sum_k\alpha_k)$. Similarly the marginal likelihood for the alternative model is given by
%\[
%\mathrm{L}_1(\boldsymbol{\alpha}^{(u)},\boldsymbol{\alpha}^{(s)}_{i}|\mathbf{n}^{(s)}_{i},\mathbf{n}^{(u)}_{i}) =\frac{\mathrm{B}(\boldsymbol{\alpha}^{(u)}+\mathbf{n}^{(u)}_{i}) \mathrm{B}(\boldsymbol{\alpha}^{(s)}+\mathbf{n}^{(s)}_{i})}{\mathrm{B}(\boldsymbol{\alpha}^{(s)})\mathrm{B}(\boldsymbol{\alpha}^{(u)})} \cdot \frac{N^{(s)}_{i}!}{\prod_k n^{(s)}_{ik}!} \frac{N^{(u)}_{i}!}{\prod_k n^{(u)}_{ik}!}.
%\]
The estimation procedures (both EM and MCMC) for the Dirichlet--multinomial are the same as for the beta-binomial model except that the number of parameters to estimate is larger. We initialize the $z_i$ in the EM algorithm with the positivity calls from the multivariate Fisher's exact test. In our experience, the performance of the EM algorithm greatly deteriorates for $K>3$, and is more dependent on the initial values and can fail to converge in many instances. Although our MCMC algorithm is slightly more computational, it does not suffer from this problem and provides a robust alternative when $K$ is large. More details are given in Appendix C of the Supplementary Material.

\subsection{Polyfunctionality in Fluidigm Single-Cell Gene Expression Data}
As a proof-of-concept, we applied our multivariate MIMOSA model to two specific genes in the Fluidigm data, namely CCR7 and GZMK. For this example $K=2$ and we have four possible subsets.
In Figure~\ref{fig:polyfunctionality} we show heatmaps of the $\log_2$ counts of cells expressing all combinations of the CCR7 and GZMK genes in unstimulated and stimulated samples, as well as for the $log_2$ of the sum of the counts of the 3 positive subsets. The number of CCR7-GZML+ cells increases upon stimulation, while the number of CCR7+GZMK- cells decreases. The typical approach to analyzing poly-functional populations from intracellular cytokine staining data by summing the counts over all polyfunctional cell populations (\textit{i.e.}, cells expressing CCF7$+$ or GZMK$+$), would not be appropriate in this case, since changes in the counts are bidirectional and their sum is near zero. The multivariate MIMOSA model tests for any difference across all polyfunctional cell subsets and detects significant differences in 12 of the 16 subjects (Figure~\ref{fig:polyfunctionality} A, dark labels), whereas the standard approach detects no differences (Figure~\ref{fig:polyfunctionality} B). Testing all combinations simultaneously is also an advantage over performing multiple univariate tests on the subject combinations, which requires multiplicity adjustment and a potential loss of power.

@@

Since the Fluidigm data set has a limited number of observations (100 cells and 16 samples), we could not look at more than two biomarkers at once. Therefore, we performed simulations in eight dimensions to assess the performance of the multivariate MIMOSA  model on the resulting 2x8 tables (Supplementary Figure 6 A,B). These results show that multivariate MIMOSA outperformed competing methods even when cell counts were small and at small effect sizes. MIMOSA also provided better FDR control than competing approaches (Supplementary Figure 6 B).

@@

\section{Discussion}
\label{s:discussion}
The development of effective statistical methods to detect differences in gene or protein expression at the single-cell level is becoming increasingly important as single-cell assays become more routine and  sequencing at the single-cell level eventually becomes practical~\citep{Ramskold:2012gj}. Current approaches, such as the t-test, $\chi^2$ test and Fisher's exact test are for the most part simplistic, and resulting inference can be quite sub-optimal, especially when the cell counts are small. Most importantly, these methods do not share information across samples, resulting in less power to detect true differences than empirical-Bayes and hierarchical modeling approaches, which are widely applied in the microarray literature~\citep{Kendziorski:2003uw,Newton2004,Smyth:2005iy,Gottardo:2006}. In addition, most of these methods are univariate in nature and inappropriate for high--dimensional, next--generation single-cell assays.

The MIMOSA model uses a mixture framework to model counts in experimental subjects across multiple conditions. Information is shared across subjects through exchangeable priors, increasing the power to detect true differences compared to competing methods, even when the underlying model assumptions are violated (Figures~\ref{fig:simulations} and Supplementary Figure 5 B). Although in some instances the FDR was not well controlled by any method tested (Supplementary Figure 1A), volcano plots of effect size vs posterior probability of response show these false positives are pre--vaccine subjects exhibiting response  to stimulation (Supplementary Figure 7). Such variability is occasionally expected due to assay failures or biological variation.

MIMOSA is fit to dichotomized data  (cells are either positive or negative). While this is the standard approach for ICS data analysis, it is believed that the magnitude (fluorescence intensity) of the signal also carries information. For single-cell gene expression data, most methods have focused on differences in the continuous part of the signal while using the discrete on/off nature for pre-filtering~\citep{Flatz:2011jb}. 
The ability of MIMOSA to identify stimulation-specific expression patterns in the Fluidigm data demonstrates not only the broader utility of the method, but demonstrates that biologically relevant signal is present in the frequency of gene expression (Figure~\ref{fig:fluidigm}). 
We have shown elsewhere that changes in both the frequency of expression and expression level are informative~{McDavid2012}, and extensions of the model that incorporate the continuous component (MFI or gene expression) are warranted in the future. 
Although we used two single-cell assay platforms as motivating examples, our MIMOSA model can be applied to any type of single-cell assay where cells are dichotomized into positive and negative subsets.

@@

Detecting differences in poly-functional cell populations is important in immunology, since it allows the identification of more homogeneous cell populations~\citep{Milush:2009bz}.
In the context of HIV, poly-functional cell populations have been shown to be correlated with long-term disease non-progression, while in the context of vaccination studies poly--functional responses have been correlated with protection from disease~\citep{Betts:2006dw,Darrah:2007ih,Precopio:2007ht}.
In the ICS data used here, the stimulation is expected to only increase the number of antigen specific cells. The univariate MIMOSA model allows us to constrain the alternative hypothesis to the case $p^{(s)} > p^{(u)}$, where the proportion of cells in the stimulated sample is strictly greater than the proportion of cells in the unstimulated sample, and is ideal for this situation.
Hence, if a specific cell subset expressing multiple biomarkers is differentially expressed, then differential expression based on the marginal cell counts should also be detectable.
As such, identifying poly-functional cytokine profiles from ICS data could be done in an iterative way.
First, univariate tests on marginal populations are performed, then specific cell subsets expressing the positive biomarkers detected are tested.
However, this iterative (univariate) approach might not be satisfactory due to the large number of combinations that need to be tested. A multivariate approach might be preferable.
In that case,  as others have pointed out, in order to have the most power to detect a true difference, the statistical test should only take into account the cytokine combinations of interest~\citep{Nason:2006dx}.

When changes are two--sided (as with Fluidigm), differences in poly-functional cell populations are not always detectable  since differences may cancel in the marginal populations (Figure~\ref{fig:polyfunctionality} B).
In this case, the use of multivariate models, as our Dirichlet-multinomial model, will be important (Figure~\ref{fig:polyfunctionality} A).
Unfortunately, the limited number of samples in the Fluidigm data prevented us from looking at co-expression involving more than two genes.
In the case of more than two biomarkers, the number of parameters for our model is $2^{K+1}+1$, which is large even for moderate values of K. Even for  $K=4$, we would need a large number of subjects and cells to properly estimate the 33 parameters. A solution would be to explore alternative model parameterizations that could be used to reduce the number of required parameters.
For example, one could assume that the hyper-parameters are constant across biomarker combinations, \textit{i.e.}, $\alpha^{\{s,u\}}_{k}=\alpha^{\{s,u\}}$ for all $k$, and the number of parameters would be reduced to $3$ for any $K$.
As attractive as this might sound, such a model would be unrealistic given that certain stimulations are known to induce expression of certain biomarkers more than others.
More exploratory work will need to be done in this area once high dimensional single-cell level data with large number of samples become available.
